---
title: "Nearest Neighbor Search"
author: "David Oliver"
date: "2/4/2020"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(babynames)
library(plotly)
library(magrittr)

```

# Framing the Question

Given a set $S$ of points in a space $M$ and a query point $q \in M$, find the closest point 
in $S$ to $q$. Most commonly, $M$ is taken to be the d-dimensional vector space where 
dissimilarity is measured using some distance metric.

## Basic Scenario

Let's start by defining a set of points, $S$, in a 2-dimensional space, $M$. 

```{r, Initial Data}

S <- babynames$name %>% unique() %>% sample(100, replace = F) %>% 
     data.frame(Names = ., X = rnorm(100), Y = rnorm(100))

plot_ly(S, x = ~X, y = ~Y, 
        type = "scatter", mode = "markers", 
        name = "babies", hoverinfo = "text", text = ~Names)

```

Now define a new point, $q$, also in 2-dimensions.  

```{r}

q <- data.frame(Names = "q", X = rnorm(1), Y = rnorm(1)) 

rbind.data.frame(S, q) %>% add_column(., source = c(rep(TRUE, 100), FALSE)) %>% 
    plot_ly(x = ~X, y = ~Y, type = "scatter", mode = "markers", name = "babies",
        hoverinfo = "text", text = ~Names, 
        color = ~source, colors = c("#d73027", "#1a9850"))

```

That's it, we're all set up. Now we want to find the nearest neighbor to the point $q$. 

Determining the nearest neighbor means that we need to find some distance, or more generally 
a dissimilarity, metric on which to judge nearness.

The choice of metrics is dependant on your problem and there is no clear method for making
the decision other than rationale relating to the specifics of the data and the question.

In the mean time, if you are desparate for a place to start and all the options have given
you analysis paralysis, try [GUSTA ME](https://sites.google.com/site/mb3gustame/wizards/-dis-similarity-wizard)'s 
(Dis)similarity wizard as a useful guide for where to start.

<!-- A good guideline to follow regarding the choice of distance/dissimilarity metrics would be -->
<!-- that if the results of the analysis change dramatically with a change in dissimilarity metric,  -->
<!-- identifying the reason for the variability would benefit the analysis more than choosing the -->
<!-- "correct" metric. -->

That being said, for this example, there really isn't any reason to get fancy with the 
metric yet, so we will stick with euclidean distance. The Euclidean distance in 2D space 
between any two points $p$ and $q$ is $d=\sqrt{(x_{p}-x_{q})^2+(y_{p}-y_{q})^2}$

```{r}

# if p and q are two n dimensionsal points, then the euclidean distance is
euclidist <- function(p, q){
    stopifnot(length(p) == length(q))
    sqrt(sum((p - q)^2))
}

S$euclid_dist <- apply(S[,2:3], 1, euclidist, q = q[,2:3])

S[which.min(S$euclid_dist),]

plot_ly(S, x = ~X, y = ~Y, type = "scatter", mode = "markers", name = "babies",
        hoverinfo = "text", text = ~Names) %>% 
    add_segments(., x = ~X, xend = q$X, y = ~Y, yend = q$Y,
                 line = list(color="#000000", width=0.5, dash="solid"))

```

The nearest neighbor to q in space M from set S is `r min(S$euclid_dist)` units away and is 
named `r S$Names[which.min(S$euclid_dist)]`.

# Extension of Nearest Neighbors Search

An interesting property of nearest neighbors (really it is a property of distance metrics) 
is that a given point $p$ may have nearest neighbor $q$, but $p$ is not necessarily $q$'s 
nearest neighbor. 

At the moment I'm not sure what the usefulness of this property is, but let's try to find
the nearest neighbor for each point in our dataset (we'll omit our initial "q" point). 
In order to gain some computational efficiency we'll just use `dist` from the `stats` package.

```{r, fig.height=7}

# calculate the distance between each pair of points
distMat <- 
    dist(S[, c("X", "Y")], method = "euclidean", upper = T) %>% 
    as.matrix %>% { diag(.) <- NA }

# the nearest neighbor of each point is the minimum value in each row or column of the 
# square distance matrix (with diag = NA)
minDist <- apply(distMat, 1, which.min)

# collecting the target and destination nodes
S$NNFrom <- as.numeric(names(minDist))
S$NNTo <- minDist

# get X-Y values for both origin (X0,Y0) and target (X1,Y1) nodes.
SNet <- data.frame(Name0 = S$Names, X0 = S$X, Y0 = S$Y, X1 = S$X[S$NNTo], Y1 = S$Y[S$NNTo])

plot_ly(SNet, x = ~X0, y = ~Y0, type = "scatter", mode = "markers",
        marker = list(size = 10, color = "black"), hoverinfo = "text",
        text = ~Name0) %>% 
    layout(xaxis = list(title="", zeroline = F, showticklabels = F, showgrid = F), 
           yaxis = list(title="", zeroline = F, showticklabels = F, showgrid = F)) %>% 
    add_annotations(x = ~X1, ax = ~X0,
                    y = ~Y1, ay = ~Y0, 
                    xref = "x", yref = "y",
                    axref = "x", ayref = "y",
                    text = "", arrowhead = 2,
                    showarrow = T)

```

From this directed graph, we can see that nearest neighbors are not always recipricol.




The simplest extension of the nearest neighbor search is the K-NN variation. That is,
given the distance (dissimilarity) from a point to it's neighbors find the K closest 
neighbors. 

For my purposes I want to explore k-NN and s-NN variants of the nearest neighbor search.







It seems that maybe this is important for nearest-neighbor chain algorithms. In addition,
this concept might be important for S-NN and M-NN approaches. 